---
title: "2022-23 NBA Scoring"
author: "Conor Cassidy"
date: "6 January 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Offensive Excellence...
This is just a short investigation into the scoring in the 2022/23 NBA season so far. It comes after a conversation with a friend in which we were discussing how the numbers that some players have been putting up recentely are ridiculous (e.g. D. Mitchell 71 point career high, L. Doncic's 60/20/10 game, Giannis w/ a career high etc.) and it made me question this offensive explosion... 

# Data 
from NBA website: https://www.nba.com/stats/players/traditional?TeamID=1610612749&Country=GR&PlayerPosition=F&StarterBench=Starters&DraftYear=2013&Season=2022-23

```{r echo = FALSE, include=FALSE}
library(robotstxt)
library(tidyverse)
library(robotstxt)
library(dplyr)
library(rvest)
paths_allowed("https://www.nba.com/stats/players/traditional?TeamID=1610612749&Country=GR&PlayerPosition=F&StarterBench=Starters&DraftYear=2013&Season=2022-23")

page <- read_html("https://www.nba.com/stats/players/traditional?TeamID=1610612749&Country=GR&PlayerPosition=F&StarterBench=Starters&DraftYear=2013&Season=2022-23")
page
typeof(page)
class(page)

stats <- page %>%
  html_nodes(".Crom_container__C45Ti , :nth-child(30), .Crom_body__UYOcU :nth-child(29), .Crom_body__UYOcU :nth-child(28), .Crom_body__UYOcU :nth-child(27), .Crom_body__UYOcU :nth-child(18), .Crom_body__UYOcU :nth-child(17), .Crom_body__UYOcU :nth-child(16), .Crom_body__UYOcU :nth-child(15), .Crom_body__UYOcU :nth-child(12), .StatEventLink_sel__pAwmA, .Crom_body__UYOcU :nth-child(9), .Crom_body__UYOcU :nth-child(8), .Crom_body__UYOcU :nth-child(5), .Crom_body__UYOcU :nth-child(6), .Crom_body__UYOcU :nth-child(7), td:nth-child(4)") %>%
  html_text() 
stats
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
