---
title: "Are Films Getting Worse?"
author: "Conor Cassidy"
date: "07-01-2023"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Cinematic Slump...
I recently went to the cinema with a friend to see the new Black Panther film (I wouldn't recommend...) which sparked an interesting conversation afterwards - are films getting worse? We were trying to rack our brains of films that have came out in the last few years that were among our favourites. After we struggled to think of many, we took to IMDb to check if we were just being forgetful or if there genuinely were not that many great films released recently. I decided to do a little analysis of the Top 250 films, according to IMDb, over the years to see if they are indeed getting worse. The list can be found here via the following url: https://www.imdb.com/chart/top/.

## The Data
Feel free to skip over this part - it just explains the technicalities of how I scraped the data for this analysis. It was obtained from the IMDb website's Top 250 Films List - given in url given above. The code for this is shown in the .Rmd file on the Github page. For the most part, I done it using the tools from the rvest package. One of the main reasons I chose to undertake this project was to practice some data scraping. 

```{r include = FALSE}
library(tibble)
library(tidyverse)
library(robotstxt)
library(dplyr)
library(rvest)
library(stringr)
```


```{r include = FALSE}
page <- read_html("https://www.imdb.com/chart/top/")

paths_allowed("https://www.imdb.com/chart/top/")

typeof(page)
class(page)

titles <- page %>%
  html_nodes(".titleColumn a") %>%
  html_text()

years <- page %>%
  html_nodes(".secondaryInfo") %>%
  html_text() %>%
  str_remove("\\(") %>%
  str_remove("\\)") %>%
  as.numeric()

ratings <- page %>% 
  html_nodes("strong") %>%
  html_text() %>%
  as.numeric()

imdb_top_250 <- tibble(
  title = titles,
  year = years,
  rating = ratings
)

imdb_top_250 <- imdb_top_250 %>%
  mutate(rank = 1:nrow(imdb_top_250)) %>%
  relocate(rank)

imdb_top_250$decade <- as.factor(ifelse(imdb_top_250$year < 1930, '1920s',
                                 ifelse(imdb_top_250$year %in% c(1930:1939), '1930s',
                                 ifelse(imdb_top_250$year %in% c(1940:1949), '1940s',
                                 ifelse(imdb_top_250$year %in% c(1950:1959), '1950s',
                                 ifelse(imdb_top_250$year %in% c(1960:1969), '1960s',
                                 ifelse(imdb_top_250$year %in% c(1970:1979), '1970s',
                                 ifelse(imdb_top_250$year %in% c(1980:1989), '1980s',
                                 ifelse(imdb_top_250$year %in% c(1990:1999), '1990s',
                                 ifelse(imdb_top_250$year %in% c(2000:2009), '2000s',
                                 ifelse(imdb_top_250$year %in% c(2010:2019), '2010s', '2020s')))))))))))
```

## Which years produced the best films?

```{r echo = FALSE}
top_years <- imdb_top_250 %>%
  count(year, sort = TRUE) %>% 
  rename("Year" = year, "Number of Films in the Top 250" = n) %>% 
  top_n(10) %>% 
  suppressMessages()
knitr::kable(top_years)

# by decades
ggplot(imdb_top_250, aes(x=decade, fill = decade)) +
  geom_bar(stat = "count", fill = "light blue") + 
  labs(
    x = "Decade",
    y = "Number of entries",
    title = "Number of Entries into the IMDb Top 250 Films Over the Decades") +
  theme(legend.position = "none") +
  theme_bw() 

# by years
ggplot(imdb_top_250, aes(x= years, fill = years)) + 
  geom_bar(stat = "count", fill = "light blue") +
  labs(
    x = "Year",
    y = "Number of entries",
    title = "Number of Entries into the IMDb Top 250 Films Over the Years") +
  theme(legend.position = "none") +
  scale_x_continuous(breaks=seq(1920,2025,by=10)) +
  theme_bw()
```

We can also calculate the average number of films in the top 250 from each year: $$250/(2022-1921) \approx 2.5.$$ As a result, we can say that any year with 3 or more films in the top 250 is a pretty good year for films. Likewise, if a year has no film or only 1 film in the top 250, it was a fairly bad year for films. 

We can see that the mid-to-late 90s and early-mid 2000s were great years for films - contributing many of the top 250 films. Other great years include 1957, 2019, and 1975.

## The Verdict
Taking all of the analysis into consideration, I think it's fair to conclude that there hasn't been a 'good' year for films since 2019. The last time it wasn't a 'good' year for films before this was over 20 years ago - in 1996. So yes, me and my friend seem vindicated in the view that the cinema seems to be in a bit of a slump at the minute...


## RECOUNT!!! 

After different friend of mine read this report thus far, they disagreed. They proceeded to list of some recent films that they thought were great and the report was a sham. I couldn't accept them sullying this technical report with such slander! So I decided to augment this report with a further supplementary investigation. How could this be done? I used a different source for the Top 250 All Time Films! I used another website, metacritic, find it here if you wish to browse https://www.metacritic.com/browse/movies/score/metascore/all/filtered?sort=desc. Again, like in the previous section of IMDb data, I scraped the data from the metacritic website. If you're interested in how this was done, please feel free to look at the code in the .Rmd file of the Github page. 


```{r include = FALSE}
page_2 <- read_html("https://www.metacritic.com/browse/movies/score/metascore/all/filtered?sort=desc")

paths_allowed("https://www.metacritic.com/browse/movies/score/metascore/all/filtered?sort=desc")

typeof(page_2)
class(page_2)

# 1st 100 films
titles2 <- page_2 %>%
  html_nodes(".menu , h3") %>%
  html_text()
titles2
titles2 <- titles2[- c(1:13)]

# 2nd 100 films
page_3 <- read_html("https://www.metacritic.com/browse/movies/score/metascore/all/filtered?sort=desc&page=1")

# 3rd 100 films
page_4 <- read_html("https://www.metacritic.com/browse/movies/score/metascore/all/filtered?sort=desc&page=2")

# 100-200

somet <- page_3 %>%
  html_nodes(".menu , h3") %>%
  html_text()
somet <- somet[- c(1:13)]

somet1 <- page_4 %>%
  html_nodes(".menu , h3") %>%
  html_text()
somet1 <- somet1[- c(1:13)]
somet1 <- somet1[- c(51:100)]
somet1

# now top 250
titles2 <- append(titles2, somet)
titles2 <- append(titles2, somet1)
titles2

# now for the years
years2 <- page_2 %>%
  html_nodes(".clamp-details span:nth-child(1)") %>%
  html_text() 
years2

years2 <- lapply(years2, function(x) str_sub(x, - 4 ,- 1 ))
years2 <- as.character(years2)

years_200 <- page_3 %>%
  html_nodes(".clamp-details span:nth-child(1)") %>%
  html_text() 
years_200

years_200 <- lapply(years_200, function(x) str_sub(x, - 4 ,- 1 ))
years_200 <- as.character(years_200)

years_300 <- page_4 %>%
  html_nodes(".clamp-details span:nth-child(1)") %>%
  html_text() 
years_300

years_300 <- years_300[- c(51:100)]
years_300 <- lapply(years_300, function(x) str_sub(x, - 4 ,- 1 ))
years_300 <- as.character(years_300)

years2 <- append(years2, years_200)
years2 <- append(years2, years_300)

# finally the ratings too
ratings2 <- page_2 %>% 
  html_nodes(".positive") %>%
  html_text() %>%
  as.numeric()
ratings2


meta_250 <- tibble(
  title = titles2,
  year = years2)

meta_250$decade <- as.factor(ifelse(meta_250$year < 1930, '1920s',
                                 ifelse(meta_250$year %in% c(1930:1939), '1930s',
                                 ifelse(meta_250$year %in% c(1940:1949), '1940s',
                                 ifelse(meta_250$year %in% c(1950:1959), '1950s',
                                 ifelse(meta_250$year %in% c(1960:1969), '1960s',
                                 ifelse(meta_250$year %in% c(1970:1979), '1970s',
                                 ifelse(meta_250$year %in% c(1980:1989), '1980s',
                                 ifelse(meta_250$year %in% c(1990:1999), '1990s',
                                 ifelse(meta_250$year %in% c(2000:2009), '2000s',
                                 ifelse(meta_250$year %in% c(2010:2019), '2010s', '2020s')))))))))))
```


Below, I perform the same analysis as for the IMDb data but now using the metacritic data. 

```{r meta, echo = FALSE}
top_years2 <- meta_250 %>%
  count(year, sort = TRUE) %>% 
  rename("Year" = year, "Number of Films in the Top 250" = n) %>% 
  top_n(10) %>% 
  suppressMessages()
knitr::kable(top_years2)

# by decades
ggplot(meta_250, aes(x=decade, fill = decade)) +
  geom_bar(stat = "count", fill = "light blue") + 
  labs(
    x = "Decade",
    y = "Number of entries",
    title = "Number of Entries into the Metacritic Top 250 Films Over the Decades") +
  theme(legend.position = "none") +
  theme_bw() 

# by years
ggplot(meta_250, aes(x= years2, fill = years2)) + 
  geom_bar(stat = "count", fill = "light blue") +
  labs(
    x = "Year",
    y = "Number of entries",
    title = "Number of Entries into the Metacritic Top 250 Films Over the Years") +
  theme(legend.position = "none") +
  scale_x_discrete(breaks=seq(1920,2025,by=10)) +
  theme_bw()
```

To get a better idea of how this compares to the IMDb data, I made the following plot and graph:

```{r combined, echo = FALSE}
# Plot to compare the number of films in each lists top 250 per decade
ggplot(NULL, aes(x = decade)) + 
  geom_bar(data = meta_250, aes(fill = "metacritic"), stat = "count", alpha = 0.6, show.legend = TRUE) +
  geom_bar(data = imdb_top_250, aes(fill = "IMDb"), stat = "count", alpha = 0.5, show.legend = TRUE) +
  labs(
    x = "Decade",
    y = "Number of Entries",
    title = "Number of Entries into the Top 250 Films Over the Decades") +
  scale_fill_manual(values = c("metacritic" = "yellow", "IMDb" = "blue"), labels = c("IMDb", "metacritic")) +
  scale_x_discrete(breaks = seq(1920, 2025, by = 10)) +
  theme_bw() + 
  theme(legend.title = element_blank())


# Table w/ recent 5 years comparing # in top 250 for meta & imdb
meta_recent <- meta_250 %>%
  count(year, sort = TRUE) %>% 
  rename("Year" = year, "Number of Films in the Top 250" = n) %>% 
  arrange(desc(Year)) %>% 
  slice(1:6)

imdb_recent <- imdb_top_250 %>%
  count(year, sort = TRUE) %>% 
  rename("Year" = year, "Number of Films in the Top 250" = n) %>% 
  arrange(desc(Year)) %>% 
  slice(1:6)

# Convert the "Year" column to character type in both tables
meta_recent$Year <- as.character(meta_recent$Year)
imdb_recent$Year <- as.character(imdb_recent$Year)

# Perform the left join and rename columns
combined_table <- left_join(meta_recent, imdb_recent, by = "Year") %>%
  rename(Meta = `Number of Films in the Top 250.x`,
         IMDb = `Number of Films in the Top 250.y`) %>%
  select(Year, Meta, IMDb)

knitr::kable(combined_table)
```

N.B., in the Year Column above that TBA ("To Be Announced") means that the film is not yet released. This is the case for two of the films in the metacritic top 250 list ("The Zone of Interest", and "The Delinquents"). This basically just means that they are previewed by critics at e.g. Cannes Film Festival and scored based from these. 

FIX OVERLAYED PLOT X-AXIS

It is now easier to see and definitively say that these two sources differ in their rankings of recent films. Metacritic ranks more recent films higher than IMDb does in their top 250 rankings. This can be seen in the graph above, where the two most recent decades (20s & 10s) have more top 250 films than IMDb, whereas IMDb has more top 250 films allocated in the decades of the 80s, 90s, & 00s. 

In fact, metacritic has more films in their top 250 list than IMDb in each of the last 5 years.  

So, perhaps unsurprisingly, the answer to the question "Are films getting worse?" seems to be subjective. If you are inclined to agree with me, you might just want to use IMDb to help choose which films to watch next, whereas if you are a fan of more recent cinema you may wish to use metacritic to decide which movie to watch next. 

Thank you for reading!